---
format:
  html:
    toc: true
    toc-location: left
    toc-title: "In this session:"
editor: 
  markdown: 
    wrap: 72
---

# Tutorial 9: BIOL90042 final project {#sec-tute09}

In this tutorial, we will introduce the final group project for
BIOL90042, where you will analyse several related datasets in the
context of a research program on Parkinson's Disease.

## Background

### Parkinson's Disease

Parkinson's Disease (PD) is a neurodegenerative condition caused by the
progressive loss of dopamine-producing neurons in a region of the brain
called the substantia nigra. Before these neurons die, misfolded
alpha-synuclein protein starts to build up inside them and this
build-up, along with mitochondrial dysfunction, oxidative stress and
neuroinflammation, seems to play a key role in driving the disease.

Patients with PD experience motor symptoms such as slowed movement,
tremor, poor balance, an unsteady gait and more frequent falls.
Non-motor symptoms like depression, fatigue, cognitive impairment,
psychotic symptoms and dementia are also common, and can be just as
disruptive to a patient's quality of life.

What causes PD is not fully understood. Genome-wide association studies
(GWAS) have identified many common genetic variants (SNPs) that are more
common in PD patients than in the general population. Patients having
many of these common SNPs are considered to have a high polygenic risk.
A smaller group of patients (around 10%) instead carry one of a handful
of rare, high-risk SNPs that substantially raise their chances of
developing PD. For everyone else, the disease is considered ‘sporadic’,
meaning it is not clearly tied to genetics, but might instead be linked
to long-term exposure to pesticides, heavy metals or air pollution.

PD motor symptoms can be managed in the early disease stage using the
dopamine supplement L-DOPA, however treatments to prevent neuronal cell
death are needed, and to date there is no cure for PD.

### Project scenario

In this project, your group is a team of biomedical researchers tasked
with analysing the data from WEHI’s Parkinson’s Disease research
program, which has the following four arms: 1. Population genetics 2.
RNA-seq of patient iPSC-derived neurons 3. Drug screening 4. Clinical
trial

As a group, you will need to use R to analyse data from these four arms,
and present them in both a **quarto report (`.qmd`, combining code,
plots and explanations)** and **oral presentation**. Participants are
identified by a unique donor id (`donorid` column in the datasets), and
some participants took part in multiple study arms, so you may use them
to connect the different datasets.

## Overview of the datasets

Each group will have their own simulated datasets, available to download
from canvas. They are provided to you as R data files (`.Rda`), which
you can read in using the `load()` function, or by clicking on them from
the files panel.

We will now go through the datasets and briefly explain their contents.

```{r}
#| message: false
library(tidyverse)
load("data/example_group_project_data/prs.Rda")
load("data/example_group_project_data/rnaseq.Rda")
load("data/example_group_project_data/rnaseq_gene_info.Rda")
load("data/example_group_project_data/drugscreen.Rda")
load("data/example_group_project_data/gait.Rda")
```

### Population genetics

This data is in the `prs.Rda` file.

A GWAS has been performed on Australian PD patients (cases) and healthy
controls. The top hits are summarised in the `gwas_top_hits` dataframe:

```{r}
gwas_top_hits
```

For each of these 15 loci, the reference and alternative alleles are
given, and a GWAS risk coefficient (`beta` column) has been calculated.
You will need this information to calculate polygenic risk scores (PRS)
for the patients in this study.

The genotypes of 150 PD patients and 150 healthy controls at these 15
loci were determined, and are provided in the `ca_co_genotypes`
dataframe:

```{r}
ca_co_genotypes
```

There is one column for each risk locus, and the values in these columns
represent the 'risk genotype dosage' (how many copies of the risky
allele an individual has: 0, 1 or 2).

To calculate the PRS of an individual, take the sum of each locus' risk
genotype dosage by the corresponding GWAS risk coefficient.

More information about GWAS can be found in lecture 7.

### RNA-seq of patient iPSC-derived neurons

This data is in the `rnsaeq.Rda` (DGEList) and `rnaseq_gene_info.Rda`
(gene names and gene ontology data).

To compare the transcriptional profiles of PD neurons compared to
healthy controls, induced pluripotent stem cells (iPSCs) from 8 PD
patients and 8 healthy age and sex-matched controls were differentiated
into dopaminergic neurons. After two weeks in culture, RNA was isolated,
sequenced and pre-processed to generate the DGEList object, `dge`.

You are also provided with `entrez_genename_key`, a dataframe that
connects the [Entrez
IDs](https://pmc.ncbi.nlm.nih.gov/articles/PMC1761442/) of genes (as
given in `dge$genes`) to the actual gene names and description:

```{r}
entrez_genename_key
```

[Gene ontology (GO)](https://geneontology.org/) information, which
describes the broad biological functions of genes is provided in the
`gene_ontology` dataframe:

```{r}
gene_ontology
```

GO is useful for interpreting the your differential expression (DE)
results, without needing to do extensive research on all of your DE
genes.

More information about RNA-seq analysis can be found in workshop 7 (see
@sec-workshop07).

### Drug screening

This data is in the `drugscreen.Rda` file.

To identify chemical compounds that can protect neurons from cytotoxic
stress, a drug screen was performed using a 1536-well plate containing
dopaminergic neurons from a healthy donor. 192 test compounds, and 192
control compounds were each tested in technical quadruplicate (i.e.,
each compound in 4 separate wells). The compound in each cell is
described in the `plate_meta` dataframe:

```{r}
plate_meta
```

Each test compound has a control which has the same molecular weight and
molecular constituents, but different chemistry with no known biological
activity. The enzyme target and mechanism of action for each of the test
compounds is known, and described in the `compound_dict` dataframe:

```{r}
compound_dict
```

After adding the compound to the neurons, all wells were treated with
MPP, a chemical that causes oxidative stress and mitochondrial damage to
mimick PD disease biology. After 24 hours of incubation, neuronal
viability was measured in a luminescence assay. Cells that survived and
grew despite MPP exposure have the highest luminescence values, whereas
those damaged by MPP have lower luminescence values. These luminescence
values are given in relative light units (RLU) in the `plate1`
dataframe:

```{r}
plate1
```

### Clinical trial

This data is in the `gait.Rda` file.

It turns out that one of the hits from the drug screen is already
approved (for other diseases) by the Australian Therapeutic Goods
Administration. Thus, a clinical trial is designed to test the effect of
this drug on PD motor symptoms. 16 PD patients are andomly assigned to
L-DOPA (current standard of care), or the trial drug. In addition, half
of the patients in each treatment group are randomly assigned to a
supervised daily exercise routine.

All patients are given a smart watch to wear that measures their
vertical orientation (in degrees off-axis) whilst walking throughout the
day (every second, for 700 seconds). This measurement (in the `1_secs`,
`2_secs`, etc columns), is a proxy for balance and smoothness of motion
whilst walking. A measurement of 0 in indicates that the patient is
standing perfectly upright.

Note that watches are not worn during the exercise routine, just during
everyday activities.

```{r}
gait_data
```

## Research questions

You are welcome to explore the data as you wish, but **the following
questions should be answered for each dataset in your quarto report**.
Due to time constraints for the oral presentation, you will need to
choose what you think is the most interesting/relevant to present

### Population genetics

1.  Calculate a PRS for each participant. Which participant has the
    highest PRS? The lowest?
2.  Plot and describe (with descriptive statistics) the distribution of
    PRS for both cases and controls.
3.  Is there a statistically significant difference in the PRS of cases
    and controls? Ensure you check all the assumptions of the test you
    choose.

### RNA-seq of patient iPSC-derived neurons

1.  Which genes are differentially expressed between PD cases and
    healthy controls? Please also visualise your DE results with
    appropriate plot(s).
2.  Which gene ontology families do the DE genes belong to?
3.  Is the clustering of samples by gene expression consistent with
    their disease status as cases and controls?

### Drug screening

1.  Using the `geom_tile()` function, visualise the luminescence across
    the plate. Do there appear to be any spatial patterns in this data?
2.  Which compounds are statistically significantly associated with
    increased neuronal viability after MPP exposure? Ensure that you
    check your assumptions.
3.  What are the mechanisms of action and targets of the compounds you
    identify in Q2?

### Clinical trial

1.  Plot each participants' gait data and describe it using descriptive
    statistics. Which of these descriptive statistics do you think is
    most useful for capturing the overall gait smoothness and balance of
    a participant?
2.  Is there a statistically significant difference in gait between
    participants receiving L-DOPA and the new compound? Ensure that you
    check your assumptions.
3.  Does either of age or the exercise regime affect gait? If so, take
    this into consideration when investigating the relationship between
    the treatment and gait.

### Bringing it together/linking between datasets

1.  Using their unique donor IDs, how many patients appear across all
    four arms of the study?
2.  Do any variables other than gene expression distinguish the RNA-seq
    cases from controls? (Hint: try colouring points on your MDS plot by
    other information you have about the donors, from the other
    datasets)
3.  Are any of the gene targets of hit compounds from the drug screen,
    differentially expressed in RNA-seq? Which GO terms are they
    associated with and what might this tell you about the pathology of
    PD?
4.  Is there any relationship between the PRS and gait of patients in
    the clinical trial?
5.  In a paragraph, summarise your findings and suggest directions for
    future research.

## Deliverables

Each group will need to produce the following:

### Quarto report (code)

You should **begin by making an R project for your group** (see
@sec-fr-project for instructions on how to do this) where you will store
all of your data, code and any plots/outputs you save. Then, make a
quarto report (`.qmd` file) that contains **both** your analysis code
and explanations/interpretations of your findings. **You need to submit
the entire R project**, as a zip file so that we can test your code on
the data.

Tips for code/organisation:

-   You may produce multiple reports (for example one for each dataset)
    if this is easier for you to manage within your groups, as long as
    they are clearly named

-   If you find it helpful, you are welcome to work within git (see
    @sec-tute07) but this is not mandatory

-   Ensure all of your code is well-commented and the logic behind what
    you're doing is clear

-   It is expected that all of the research questions will be addressed
    in your report.

Tips for data visualisation:

-   Make sure the colour schemes are consistent (e.g. PD cases and
    healthy controls should always use the same colours across all
    plots)

-   Make sure you label your plots nicely (headings, axis labels)

-   Particularly for the oral presentation, make sure that the font size
    on the plots is big enough that we can easily read it on the slide

### Oral presentation

In the final tutorial (week 12), we will go over some tips for science
communication. In general though:

-   Due to the time limit, it won't be possible to include all of your
    results. Just choose a selection of the things you found most
    interesting and focus on presenting those clearly without rushing

-   Slides should focus primarily on graphical representations of the
    data, with summary statistics or statistical test results (e.g. p
    value) included as appropriate

-   Be prepared to answer questions about your approach to answering the
    questions (code), as well as your interpretation of the results

## A note on AI usage

You are permitted to use AI to help you with this assessment, but
remember that you are responsible for your code, so please check
anything you get from AI thoroughly! Also, if your code isn't commented
clearly, or you can't answer questions about why you chose a certain
approach (particularly over one we covered in class), you will lose
marks regardless of whether or not the output is correct, so you need to
understand the code you submit.
