---
format:
  html:
    toc: true
    toc-location: left
    toc-title: "In this session:"
editor: 
  markdown: 
    wrap: 72
---

# Tutorial 10: Applied Statistical Epidemiology with R {#sec-tute10}

Following on from our [statistics workshop](@sec-workshop06) where we
learned how to identify and formally test patterns in our data, we will
now learn how to apply this to epidemiological problems. We will start
by analysing data from a hypothetical study to answer a public health
research question.

::: {.callout-tip title="Learning Objectives"}
At the end of this session, learners should be able to:

1.  Clean and prepare epidemiological data
2.  Estimate infection prevalence (the proportion of individuals in a
    population who are infected at a given time)
3.  Test for associations between categorical variables (chi-squared
    test)
4.  Choose between parametric (t-test, ANOVA) and non-parametric
    (Wilcoxon, Kruskal-Wallis) tests for a continuous outcome
5.  Fit a linear model and check its assumptions
:::

## Background

### Epidemiology

Epidemiology is the study of diseases in populations. It is an extension
or application of statistics and investigates the patterns of disease
(e.g., infection or non-communicable) and other illnesses to help
protect the health and safety of our communities.

Epidemiologists look at patterns, trends and data related to a disease
or illness and can establish risk assessments or predictions of what to
expect next. Epidemiologists use evidence and data to make informed
decisions in public health.

### Malaria and mass drug administration

*Plasmodium falciparum* malaria remains a leading cause of childhood
mortality in sub-Saharan Africa. **Artemisinin-based combination
therapies (ACTs)** are the first-line treatment for malaria. In our
hypothetical scenario, substantial funding has recently been directed
toward administering ACTs as a **blanket strategy** rather than reactive
treatment of symptomatic cases — an approach called **mass drug
administration (MDA)**, where everyone in a population is treated
regardless of infection status.

Other control tools include **insecticide-treated bed nets (ITNs)**.

### Study scenario

Researchers surveyed a village **twice**: `survey_1`, conducted just
before an MDA program was rolled out, and `survey_2`, conducted in the
same village about 12 months later, while the program was running. The
same participants were seen in both surveys. At each visit, researchers
recorded demographic information, intervention use, current infection
status (by PCR) and parasite density.

**Research question:** *Does MDA with ACT influence the burden of
malaria?*

We break this down into:

-   Does it influence the **prevalence** of infection\*?
-   In **whom** (which age groups)?
-   Does it influence the **severity** of infection (parasite density)?

\*\*Note*: The prevalence of infection is defined as the proportion of
individuals in the population who are infected (in this case, with*
Plasmodium falciparum\*) at a given time.

::: {.callout-note title="Experimental design"}
Experimental design is a key component of epidemiology and is crucial in
the types of questions that you can answer.

In this hypothetical scenario we have multiple cross-sectional surveys.
Having surveys from *before* and *during* the intervention is much
stronger than a single cross-sectional snapshot: we can ask not just "is
prevalence lower during the program?" but "**did prevalence *fall* after
the intervention was introduced, and in whom?**" This before/after
comparison is the closest we can get to an experiment without
randomising the intervention ourselves.

Other experimental designs might include randomised controlled trials,
which are considered the "gold-standard". Participants are randomly
assigned to an experimental group (e.g. receiving the intervention) or a
control group (e.g. receiving a placebo or standard care). But these are
not always feasible in real-world settings. In this study, we're using a
quasi-experimental approach by comparing outcomes before and during the
intervention.
:::

## The dataset

`malaria_participants.csv` contains one row per participant **per
survey** with the following information:

-   `id`: identifier per person
-   `survey`: survey that the information was obtained: `"survey_1"`
    (before MDA) or `"survey_2"` (during MDA)
-   `age`: participant-identified age
-   `sex`: participant-identified sex
-   `pcr_positive`: 18S rRNA PCR result: 1 = currently infected, 0 = not
    infected
-   `parasite_density`: parasites per microlitre of blood (0 if PCR
    negative)
-   `act_use`: artemisinin-based combination therapy provided to
    participant: 1 = yes, 0 = no
-   `bednet_use`: slept under a bed net the previous two weeks: 1 = yes,
    0 = no

You can analyse the data by either looking at one survey at a time
(`filter(survey == "survey_2")`) or explicitly comparing the two
surveys.

Let's load in the `tidyverse` R package and dataset that we will be
using!

```{r}
#| message: false
library(tidyverse)
participants <- read_csv(
  "https://raw.githubusercontent.com/WEHI-Education/BIOL90042_R_Course/refs/heads/main/data/malaria_participants.csv"
)
```

## Data preparation {#sec-dataprep}

Now that you're experts in data cleaning and visualisation, work through
the following exercises to inspect the data, filter for only rows that
have observations for every column (i.e., all columns have non-missing
values) and create an age group column for categorical-based analyses.

::: {.callout-important title="Practice exercise: Inspect the data"}
Look at the data frame. How many participants are there, and how many
rows are there per survey? Check for missing values in each column of
`participants`.

<details>

<summary>Solution</summary>

```{r}
glimpse(participants)

participants %>% count(survey) # 900 rows in survey_1, 900 in survey_2
participants %>% summarise(across(everything(), ~ sum(is.na(.))))
```

Only `pcr_positive` and `parasite_density` have missing values (25
rows).

Note that here we used the `glimpse()` function which provides an
overall summary of the dataframe, just like the base R `str()` function
(which is what is shown in the environment panel).

</details>
:::

::: {.callout-important title="Practice exercise: Complete-case analysis"}
Epidemiological analyses typically use complete records: observations
with missing data on *any* variable used in the analysis are excluded.

Create a cleaned data frame containing only rows with complete data on
all variables in `participants`. How many rows were excluded, and why
might dropping them bias (or not bias) your results?

*Hint: `na.omit()` or `drop_na()`.*

<details>

<summary>Solution</summary>

```{r}
full_df <- participants
complete_df <- participants %>% na.omit()

nrow(full_df) - nrow(complete_df) # 25 rows excluded

# let's write this into our participants df for this tute
participants <- participants %>% na.omit()
```

25 rows were excluded, all because `pcr_positive` (and therefore
`parasite_density`) was missing. That is only \~1.4% of the data, and
the missingness is spread across groups rather than concentrated in one,
so dropping them is unlikely to bias the results.

*Note*: Missingness is very important to account for in epidemology.
Datasets are inherently messy, particularly when we have survey data and
things are coded from a dictionary (i.e., pcr_positive == 1, which means
"yes" and pcr_positive == 0, which means "no", but this can get out of
hand when we have more variables that have been translated into numeric
variables that do not immediately make sense to humans). If you're
interested in learning more about how to approach handling missing data,
see the chapter in [The Epidemologist R
Handbook](https://www.epirhandbook.com/en/new_pages/missing_data.html).

</details>
:::

:::: {.callout-important title="Practice exercise: Create PCR Status and Age Group Columns"}
Create two columns using `mutate()` and `case_when()` functions.

-   `pcr_status`: 1 == yes, 0 == no
-   `age_group`: levels **0–5**, **6–10**, **11–20**, and **21+**.

Note that age groups are defined from each participant's age *at that
survey*, so some participants move up a group between the two visits.

::: {.callout-warning title="Caution when turning a continuous variable into a categorical variable"}
We've created `age_group` as a categorical variable with specific bins
because in malaria endemic settings, young children typically carry the
highest infection burden and acquired immunity develops with age and
exposure. This is based on our *prior* knowledge of age effects in the
population and how this may influence the patterns of disease
prevalence, exposure and transmission.

Grouping allows us to capture epidemiologically meaningful patterns
without assuming a linear effect of age. While this makes things easier
to interpret, particularly for this research question, categorising
loses information and statistical power compared to modelling age
continuously.
:::

<details>

<summary>Solution</summary>

```{r}
participants <- participants %>%
  mutate(
    pcr_status = case_when(
      pcr_positive == 1 ~ "yes",
      TRUE ~ "no"
    ),
    age_group = case_when(
      age <= 5 ~ "0-5",
      age <= 10 ~ "6-10",
      age <= 20 ~ "11-20",
      TRUE ~ "21+"
    )
  )
```

</details>
::::

## Question 1: Does MDA influence the prevalence of infection?

Now let's put our **epidemiological thinking caps** on. We first ask:
what is the prevalence of infection in each `survey`?

```{r}
participants %>%
  # group by survey
  group_by(survey) %>%
  # prevalence = number pcr positive / total tested in group
  summarise(
    n_tested = n(),
    n_positive = sum(pcr_positive),
    prevalence = n_positive / n_tested
  )
```

Prevalence fell substantially from `survey_1` to `survey_2`. We would
like to use our $\chi^2$ test to test the null hypothesis that there is
no association between two categorical variables: PCR status and survey
(i.e., they are independent).

Note that we must check that we meet the assumptions of a $\chi^2$ test:

i.  our observations are independent of each other,
ii. the categories are mutually exclusive and
iii. counts are reasonably large.

For our dataset:

i.  each `id` is counted once per survey and participants do not
    influence each other
ii. our categories are mutually exclusive (every `id` falls into one
    survey and one PCR status)
iii. counts are large (\>100 per group)

::::::::::::::: {.callout-important title="Practice exercises"}
Try these practice questions to test your understanding!

:::::::: question
1\. Which line of code would correctly generate a contingency table to
evaluate the prevalence of infection in each survey?

::::::: choices
::: choice
`table(participants$survey)`
:::

::: {.choice .correct-choice}
`table(participants$survey, participants$pcr_status)`
:::

::: choice
`chisq.test(participants$survey, participants$pcr_status)`
:::

::: choice
`pivot_wider(participants, survey, pcr_status)`
:::
:::::::
::::::::

:::::::: question
2\. Say we've saved this contingency table as `survey_pcr_table`. Which
line of code correctly runs the $\chi^2$ test on this contingency
table?

::::::: choices
::: choice
`survey_pcr_table %>% summary()`
:::

::: {.choice .correct-choice}
`chisq.test(survey_pcr_table)`
:::

::: choice
`chisq.test(survey, pcr_status)`
:::

::: choice
`t.test(survey_pcr_table)`
:::
:::::::
::::::::

<details>

<summary>Solutions</summary>

1.  `table(participants$survey, participants$pcr_status)` builds a
    contingency table cross-tabulating the counts of every combination
    of `survey` and `pcr_status`. Note that option 3 returns a test
    result directly but skips the intermediate step of *creating* the
    table we asked for.
2.  `chisq.test()` takes the contingency table as its argument.

</details>
:::::::::::::::

```{r}
# create our contingency table
survey_pcr_table <- table(participants$survey, participants$pcr_status)

# perform chisq test
chi_sq <- chisq.test(survey_pcr_table)

# look at our expected values
chi_sq$expected
# look at our observed values (same as contingency table)
chi_sq$observed
# results
chi_sq
```

The $\chi^2$ test shows us that *p* \< 0.001, therefore there was a
statistically significant difference in the prevalence of infection
between the two surveys: infection prevalence **fell** after the MDA
program was introduced.

## Question 2: In whom? Prevalence by age group

```{r}
participants %>%
  # group by age group and survey
  group_by(survey, age_group) %>%
  # prevalence = number pcr positive / total tested in group
  summarise(
    n_tested = n(),
    n_positive = sum(pcr_positive),
    prevalence = n_positive / n_tested
  )
```

The youngest children (0–5) carry the highest prevalence in both
surveys, but also show the largest absolute decline between surveys.
Let's test whether infection prevalence is associated with age group
**within each survey**.

::::::::: {.callout-important title="Practice exercises"}
:::::::: question
1\. How can we create a dataset containing only the most recent survey
following the MDA program?

::::::: choices
::: choice
`survey_2 <- participants %>% filter(act_use == 1)`
:::

::: choice
`survey_2 <- participants`
:::

::: {.choice .correct-choice}
`survey_2 <- participants %>% filter(survey == "survey_2")`
:::

::: choice
`survey_2 <- participants %>% select(survey)`
:::
:::::::
::::::::

<details>

<summary>Solution</summary>

`survey_2 <- participants %>% filter(survey == "survey_2")` will filter
the dataset to the `"survey_2"` survey only, excluding the `"survey_1"`
baseline survey which took place before the MDA intervention.

</details>
:::::::::

```{r}
survey_1 <- participants %>% filter(survey == "survey_1")
survey_2 <- participants %>% filter(survey == "survey_2")

chisq.test(table(survey_1$age_group, survey_1$pcr_status))
chisq.test(table(survey_2$age_group, survey_2$pcr_status))
```

In both surveys, infection is significantly associated with age group
(*p* \< 0.001), with the youngest children (0–5) most affected —
consistent with the gradual development of acquired immunity with age
and exposure.

## Question 3: Does MDA influence the severity of infection?

Severity is measured by `parasite_density` **among infected
individuals** (`pcr_positive == 1`). PCR-negative participants have a
density of 0 by definition, so we exclude them from severity analyses.

```{r}
infected <- participants %>% filter(pcr_positive == 1)

# visualise the data first
ggplot(infected, aes(x = survey, y = parasite_density)) +
  geom_boxplot()
```

### Two groups: survey_1 vs survey_2 (t-test or Wilcoxon?)

A t-test assumes the outcome is approximately normally distributed
**within each group**. As in [Workshop 6](@sec-ttest), we check this
with a QQ-plot using `geom_qq()` and `geom_qq_line()`, faceted by group,
and confirm with a Shapiro-Wilk test in each group:

```{r}
ggplot(infected, aes(sample = parasite_density)) +
  geom_qq() +
  geom_qq_line() +
  # facet because we need to check normality for each group
  facet_wrap(~survey)
```

```{r}
infected %>%
  filter(survey == "survey_1") %>%
  # shapiro.test() needs a vector, so use pull()
  pull(parasite_density) %>%
  shapiro.test()

infected %>%
  filter(survey == "survey_2") %>%
  pull(parasite_density) %>%
  shapiro.test()
```

Both groups curve away from the reference line and both Shapiro-Wilk p
values are below 0.05, so parasite density is **not** normally
distributed in either survey — the non-parametric **Wilcoxon rank-sum
test** is the valid choice here. We run the t-test as well so you can
compare the outputs:

```{r}
t.test(
  parasite_density ~ survey, # compare density across surveys
  data = infected
)
wilcox.test(parasite_density ~ survey, data = infected)
```

Both tests agree: parasite density among the infected is significantly
lower in `survey_2`. When data are skewed, the Wilcoxon result is the
more defensible one to report — here the t-test happens to agree.

### Multiple groups: density across age groups (ANOVA or Kruskal-Wallis?)

The same logic extends to \>2 groups: **one-way ANOVA** is the
parametric option; **Kruskal-Wallis** is its non-parametric counterpart.

```{r}
ggplot(infected, aes(x = age_group, y = parasite_density)) +
  geom_boxplot()

# check normality within each group
ggplot(infected, aes(sample = parasite_density)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~age_group)

infected %>%
  filter(age_group == "0-5") %>%
  pull(parasite_density) %>%
  shapiro.test()
```

Normality is again violated, so we show both tests:

```{r}
aov(parasite_density ~ age_group, data = infected) %>% summary()
kruskal.test(parasite_density ~ age_group, data = infected)
```

Density declines with age — consistent with the development of acquired
immunity — and both tests support a significant difference between age
groups.

::::::::::::::: {.callout-important title="Practice exercises"}
:::::::: question
1\. If you check the normality assumption for a continuous variable in
two groups using a QQ-plot and Shapiro-Wilk test, and find that the
assumption is clearly violated in both groups, which test should you use
to compare them?

::::::: choices
::: choice
`t.test()`, since it's the standard test for comparing two groups
:::

::: {.choice .correct-choice}
`wilcox.test()`, since it doesn't assume normality
:::

::: choice
`aov()`, since it's more powerful
:::

::: choice
`chisq.test()`, since the normality assumption is violated
:::
:::::::
::::::::

:::::::: question
2\. You want to compare `parasite_density` across the four age groups,
but a QQ-plot shows the data is clearly non-normal in at least one
group. Which function should you use?

::::::: choices
::: choice
`aov()`
:::

::: choice
`t.test()`
:::

::: {.choice .correct-choice}
`kruskal.test()`
:::

::: choice
`wilcox.test()`
:::
:::::::
::::::::

<details>

<summary>Solutions</summary>

1.  When normality is clearly violated, `wilcox.test()` (the Wilcoxon
    rank-sum/Mann-Whitney test) is the appropriate choice, since it
    doesn't rely on the normality assumption that `t.test()` requires.
2.  `kruskal.test()` is the non-parametric equivalent of a one-way
    ANOVA, and is the appropriate choice when comparing three or more
    groups and the normality assumption is violated.

</details>
:::::::::::::::

## Question 4: A linear model

Comparing groups one at a time is limited. A **linear model** lets us
test a formal hypothesis while adjusting for other variables. Our
hypothesis:

*Among infected individuals, log parasite density declines with age and
is lower during MDA (survey_2).*

We model `log(parasite_density)` because parasite densities span orders
of magnitude and are right-skewed; logging makes the relationship closer
to linear and the residuals closer to normal.

::::::::::::::: {.callout-important title="Practice exercises"}
:::::::: question
1\. How can we create a linear regression to examine the relationship
between age and log parasite density in our `infected` dataset,
adjusting for survey?

::::::: choices
::: choice
`lm(age ~ log(parasite_density) + survey, data = infected)`
:::

::: {.choice .correct-choice}
`lm(log(parasite_density) ~ age + survey, data = infected)`
:::

::: choice
`lm(log(parasite_density) + age + survey, data = infected)`
:::

::: choice
`glm(log(parasite_density) ~ age + survey, data = infected, family = "binomial")`
:::
:::::::
::::::::

:::::::: question
2\. What are the key assumptions we must check when performing a linear
regression?

::::::: choices
::: {.choice .correct-choice}
Linearity (relationship between predictor and outcome is a straight
line), Independence (residuals are independent of each other),
Homoscedasticity (residuals have roughly constant variance across a
range of fitted values) and Normality (residuals are normally
distributed).
:::

::: choice
The residuals must follow a binomial distribution, the sample size must
be at least 1000, and all variables must be measured in the log scale.
:::

::: choice
Linearity (relationship follows a straight line), Random sampling (data
collected randomly), Normality (the outcome variable is normally
distributed) and No outliers (extreme values must be removed before
fitting).
:::

::: choice
Independence (residuals are independent), Homoscedasticity (constant
residual variance), Additivity (all predictor effects are
multiplicative) and Stationarity (mean and variance are constant over
time).
:::
:::::::
::::::::

<details>

<summary>Solutions</summary>

1.  `lm(log(parasite_density) ~ age + survey, data = infected)` fits a
    linear regression of log density on age, adjusting for survey.
    Remember the formula format is `outcome ~ predictors`. Option 4 is
    incorrect: `family = "binomial"` is for binary outcomes, not a
    continuous one.
2.  A. Correct: these are the four key assumptions of linear
    regression. B. Incorrect: the residual distribution should be normal
    (not binomial), there is no minimum sample size and the scale of
    measurement is irrelevant. C. Incorrect: it is the **residuals**
    that must be normally distributed, not the outcome variable itself;
    outliers do not need to be removed before fitting. D. Incorrect:
    multiplicative effects and stationarity are not linear regression
    assumptions.

</details>
:::::::::::::::

```{r}
model_density <- lm(log(parasite_density) ~ age + survey, data = infected)
summary(model_density)
```

Interpretation: each coefficient is a change in log-density.
Exponentiating gives multiplicative effects — e.g. `exp(coef)` for
`surveysurvey_2` is the ratio of expected density in survey_2 vs
survey_1 at the same age.

```{r}
exp(coef(model_density))
```

The p-values are very small (p \< 2e-16), telling us that log parasite
density decreases with participant age and from survey 1 to survey 2.
When we exponentiating the coefficients, we see that each additional
year of age is associated with a \~4% decrease in parasite density
(1-0.96 = 0.04 or 4%), and density in survey 2 is about 40% of that in
survey 1 at the same age (0.41), consistent with an effect of the MDA
program.

:::{.callout-caution title="What if we had used `age_group` instead of `age`?"}

`age_group` is categorical (0-5, 6-10, 11-20, 21+), so
`lm(log(parasite_density) ~ age_group + survey)` would estimate
a separate effect for each age group relative to a reference
group (e.g. 0-5), rather than a single per-year slope.

This matters for two reasons. First, the interpretation changes:
instead of "each additional year of age is associated with a ~4%
decrease in density", coefficients become comparisons between
groups (e.g. "mean density in the 21+ group is X% of that in the
0-5 group"), and we can no longer talk about a rate of change
with age.

Second, the assumptions change: the continuous model assumes a
*linear* relationship between age and log-density across the
whole 1-71 year range, whereas the grouped model makes no
assumption about the shape of the age relationship — but it
throws away information by treating everyone within a group as
identical and uses more degrees of freedom. If the true
relationship is roughly linear, the continuous model is more
powerful; if it is strongly non-linear, the grouped (or a more
flexible) model may describe it better.
:::

### Assumption checking

Linear models assume linearity, independence, homoscedasticity (constant
residual variance) and normally distributed **residuals**. We check the
residuals graphically, using the same `geom_qq()` + `geom_qq_line()`
approach as in [Workshop 6](@sec-ttest):

```{r}
# residuals vs fitted: want no pattern, constant spread
ggplot(
  tibble(fitted = fitted(model_density), resid = residuals(model_density)),
  aes(x = fitted, y = resid)
) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed")

# QQ-plot of residuals: points should sit along the reference line
ggplot(tibble(resid = residuals(model_density)), aes(sample = resid)) +
  geom_qq() +
  geom_qq_line()
```

We can also confirm the normality of the residuals with a Shapiro-Wilk
test:

```{r}
shapiro.test(residuals(model_density))
```

**What do these checks tell us about our model?**

-   **Residuals vs fitted plot:** the points are scattered randomly
    around the dashed zero line with no obvious pattern and roughly
    constant vertical spread across the range of fitted values. This
    means the **linearity** and **homoscedasticity** assumptions are
    reasonably met — the model captures the systematic structure in the
    data, and the size of the errors doesn't grow or shrink with the
    fitted values. (If we instead saw a curved pattern, that would
    suggest a non-linear relationship we failed to capture; a "funnel"
    shape would suggest heteroscedasticity.)
-   **QQ-plot:** the points sit closely along the diagonal reference
    line, with only slight deviations at the extreme lower and upper
    tails. This indicates the residuals are **approximately normally
    distributed**, with a very small number of observations slightly
    further from the mean than a perfect normal distribution would
    predict.
-   **Shapiro-Wilk test:** the p value is \~0.18, which is above 0.05,
    so we **fail to reject the null hypothesis** that the residuals are
    normally distributed. This agrees with what we see in the QQ-plot:
    both checks support the normality assumption.

Taken together, the assumption checks support our log-transformed linear
model: the residuals are unstructured, have constant variance, and are
close enough to normal that the p values and confidence intervals from
`summary(model_density)` are trustworthy. Had the plots shown clear
problems instead (curved tails in the QQ-plot, a funnel in the
residual-vs-fitted plot), common remedies would be transforming the
outcome (as we did here with the log), adding missing predictors, or
falling back on the non-parametric tests we used in Question 3.

## Summary

Today we've recapped our statistics foundations and applied them to
epidemiological questions - questions that relate to uncovering the
patterns of disease or illnesses in populations.

For our hypothetical malaria dataset, we discovered that:

-   MDA with ACT was associated with a **large fall in infection
    prevalence** (chi-squared test), across all age groups but most
    markedly in children 0–5.
-   Among the infected, **parasite density was lower during MDA**
    (t-test / Wilcoxon) and **declined with age** (ANOVA /
    Kruskal-Wallis).
-   A linear model on log density confirmed both effects jointly, and
    assumption checks (residual plots, QQ-plot) supported the
    log-transformed fit.

For the practical component, we learned and revised that:

-   What the prevalence of infection is and how to calculate it
-   What variables we may need to consider in epidemiological analyses
    that may influence the spread of disease
-   When to choose parametric tests (t-tests, ANOVA), i.e., when
    residuals are roughly normal with constant variance
-   When to choose non-parametric tests (Wilcoxon/Kruskal-Wallis) i.e.,
    when the outcome is skewed and transformations don't help
-   When it might be best to use a linear regression with categorical or
    continuous predictors

If you'd like to learn more, look at the [further reading](@sec-fr10)
for this week, which includes links to Epidemiology in R resources.

You can download the handout from our tutorial session here:

{{< downloadthis resources/assign/tutorial_10_handout.pdf dname="tutorial_10_handout" label="Download the handout for tutorial 10" icon="file-earmark-arrow-down" type="primary" class="data-button" id="summarysheet" >}}
