---
format:
  html:
    toc: true
    toc-location: left
    toc-title: "In this session:"
editor: 
  markdown: 
    wrap: 72
---

# Week 10: Further Reading {#sec-fr10}

In this week's further reading, we will cover aspects of epidemiological analysis in R that didn't make it into tutorial 10. Here, we will be exploring epidemiological modelling using the `deSolve` R package which you can read about in comprehensive detail on the [deSolve Website](http://desolve.r-forge.r-project.org/). 

::: {.callout-tip title="Learning Objectives"}
At the end of this session, learners should be able to:

Build and calibrate an SIS compartmental model with the `deSolve` R package and use it to evaluate an intervention hypothesis
:::

We will be using the `deSolve` package which you will need to install:

```{r}
#| eval: false
# install the new package
install.packages("deSolve")
```

Once you've have it installed, don't forget to load it with the `library()` function:

```{r}
library(deSolve)
library(tidyverse)
```

```{r}
#| include: false
# Recreate the cleaned data from Tutorial 10
participants <- read_csv(
  "https://raw.githubusercontent.com/WEHI-Education/BIOL90042_R_Course/refs/heads/main/data/malaria_participants.csv"
) %>%
  na.omit() %>%
  mutate(
    age_group = case_when(
      age <= 5 ~ "0-5",
      age <= 10 ~ "6-10",
      age <= 20 ~ "11-20",
      TRUE ~ "21+"
    )
  )

during <- participants %>% filter(survey == "during")
```

## Compartmental modelling

So far we have statistical evidence that SMC use is associated with lower infection prevalence in 0–5 year olds, and that prevalence fell between the surveys specifically in that group. Now we ask a **mechanistic** question: **can a simple model of transmission, with an SMC-like intervention, reproduce the pattern we observed?**

### Background: the SIS model

Epidemiologists describe infectious diseases with **compartmental models**: every person sits in a compartment, and over time people move between compartments. Malaria does not confer lasting immunity — people can be reinfected after they recover — so the natural model has just two compartments:

-   **S** (susceptible): not currently infected
-   **I** (infected): currently infected

Over time, susceptibles become infected at rate $\lambda$ (the **force of infection**) and infected people recover at rate $r$. This is called an **SIS model**. The standard way to simulate one in R is the `deSolve` package, which takes care of stepping the model through time for us. A `deSolve` model always has four ingredients:

1.  a **function** describing how the compartments change,
2.  the **starting values** of each compartment,
3.  the **parameters** (here, $\lambda$ and $r$),
4.  the **time points** to simulate.

```{r}
#| message: false
# How the compartments change
sis_model <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # susceptibles: lost to infection, gained from recovery
    dS <- -lambda * S + r * I
    # infected: gained from infection, lost to recovery
    dI <- lambda * S - r * I
    # create a list with our dS and dI states
    list(c(dS, dI))
  })
}

# Starting values, parameters, and times
initial_state <- c(S = 0.99, I = 0.01)
parameters <- c(lambda = 0.15, r = 1 / 6) # per month; infections last ~6 months
times <- seq(0, 120, by = 1) # 10 years, monthly
```

We simulate with the `ode()` function and plot the result:

```{r}
output <- ode(
  y = initial_state,
  times = times,
  func = sis_model,
  parms = parameters
)
output <- as.data.frame(output)

ggplot(output, aes(x = time, y = I)) +
  geom_line() +
  labs(x = "Time (months)", y = "Prevalence of infection")
```

Notice that prevalence rises and then **settles at a steady level** (called the *equilibrium*), where new infections exactly balance recoveries. At equilibrium:

$$
\text{prevalence} = \frac{\lambda}{\lambda + r}
$$

This gives us a direct link between the model and our data: **if we know the prevalence, we can work out the force of infection.**

### Calibrate the model to the data

The 6–10 age group was too old for SMC, so their prevalence during the intervention is our best picture of "no intervention" conditions. (You could also check: was their prevalence similar in the *before* survey? If so, treating it as a stable background is reasonable.)

1.  Using the prevalence you observed in the 6–10 group during the intervention and $r = 1/6$ per month, rearrange the equilibrium formula above to solve for $\lambda$.
2.  Update `parameters` with your value and re-run the simulation. Does the model settle at the prevalence you observed?

```{r}
# prevalence in 6-10 year olds during the intervention
prev_6_10 <- during %>%
  filter(age_group == "6-10") %>%
  summarise(mean(pcr_positive)) %>%
  pull()

# rearranging prevalence = lambda / (lambda + r) gives:
r <- 1 / 6
lambda <- prev_6_10 * r / (1 - prev_6_10)
lambda # about 0.13 per month

parameters <- c(lambda = lambda, r = r)
output_calibrated <- ode(initial_state, times, sis_model, parameters) %>%
  as.data.frame()

ggplot(output_calibrated, aes(x = time, y = I)) +
  geom_line() +
  geom_hline(yintercept = prev_6_10, linetype = "dashed") +
  labs(x = "Time (months)", y = "Prevalence of infection")
# the curve settles right on the observed prevalence (~0.44)
```

### Add SMC as an intervention

Now model the 0–5 age group, where about 60% of children received SMC. Assume a protected child has their force of infection reduced by about 75% (this matches the SMC effect in your linear model: an SMC user's prevalence was roughly three quarters lower than a non-user's).

Rather than adding new compartments, we can capture the intervention by simply **averaging the force of infection across the population**:

-   40% of children are unprotected, with force of infection $\lambda$
-   60% are protected, with force of infection $0.25 \times \lambda$

Modify your model by replacing `lambda` in `parameters` with this coverage-weighted average. Re-run the simulation.

1.  What prevalence does the model settle at now?
2.  How does this compare to the prevalence you actually observed in the 0–5 age group?
3.  Plot the with-SMC and without-SMC simulations on the same axes (e.g. save both outputs and combine them with `bind_rows()`). What do you conclude?

```{r}
coverage <- 0.60
efficacy <- 0.75
lambda_smc <- (1 - coverage) * lambda + coverage * (1 - efficacy) * lambda

output_smc <- ode(
  initial_state,
  times,
  sis_model,
  parms = c(lambda = lambda_smc, r = r)
) %>%
  as.data.frame()

both <- bind_rows(
  "No SMC" = output_calibrated,
  "SMC (60% coverage)" = output_smc,
  .id = "scenario"
)

ggplot(both, aes(x = time, y = I, colour = scenario)) +
  geom_line() +
  labs(x = "Time (months)", y = "Prevalence of infection", colour = "")

# 1. with SMC, prevalence settles at about 0.30
# 2. observed prevalence in 0-5 year olds during the program was ~0.28 — a close match!
```

*Optional challenge: instead of the weighted-average shortcut, add separate protected compartments ($P$ for protected-susceptible and $I_P$ for protected-infected) to `sis_model`, with children moving into protection when they receive SMC. Does it give the same answer?*

### Interpreting the model

1.  Does the model without SMC reproduce the prevalence in the 6–10 group? Does the model with SMC reproduce the prevalence in the 0–5 group during the intervention? If the match is only approximate, what might explain the difference?
2.  This model treats each age group as a separate little world. In reality infected adults and older children can transmit to young children. Would accounting for that make the SMC effect on 0–5 year olds look bigger or smaller than in your simple model?
3.  Our two surveys were only 12 months apart. Look at your simulation: how long does prevalence take to approach equilibrium after the force of infection changes? What does that imply about whether the during-survey prevalence in 0–5 year olds should be read as the *new equilibrium* or as a population still part-way through its decline?

## Summary

In this futher reading section you have built an SIS compartmental model with `deSolve`, calibrated its force of infection from observed prevalence, and shown that an SMC-like intervention quantitatively reproduces the data. 
